A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-09-24 paper/paper.tex
paper: insert fulltab:coh too
paper: rewrite Our construction; other minor prose fixes

\section{Introduction}
\subsection*{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}
A family of sets can commonly be represented in two ways: as a family properly speaking, indexed over a given set $S$ or as a set $T$ together with a map specifying for each element of the set $T$ on which index in $S$ it depends. In the first case, we call it an \emph{indexed} presentation. In the latter case, the set associated to a given index is the fiber, i.e. inverse image, of the index under the map, so we call it a \emph{fibered} presentation. The two presentations are equivalent and the equivalence can be phrased concisely in the language of homotopy type theory~\cite{hottbook} as a fibered/indexed correspondence\footnote{At a discussion on the Zulip channel for the HoTT community, alternative names fibration/family correspondence and unbundled/bundled correspondence were proposed, among others. In our experience, the opposition fibered/indexed is spontaneously understood by many listeners. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories. The most elementary instance of the correspondence, with $\Type$ instead of $\U$, that is $(\Sigma S: \Type. (S \rightarrow T)) \simeq (T \rightarrow \Type)$, is sometimes called ``Grothendieck construction for dummies'' (its proof requires univalence, see~\cite{hottbook}).}
\begin{equation*}
  \mbox{(fibered)}\qquad(\Sigma S: \U. (S \rightarrow T)) ~\simeq~ (T \rightarrow \U)\qquad \mbox{(indexed)}
\end{equation*}
where $\U$ represents in homotopy type theory the subset of types with a unique proof of equality (within some universe).

A \emph{presheaf} on an category is a family of sets connected with maps according to the morphisms of the category. As such, it lives on the indexed side of the correspondence, opposing on the fibered side to what is called \emph{discrete Grothendieck fibrations}. However, there are situations where a presheaf $F$ can also be seen as living on the fibered side of the correspondence. Indeed, when the indexing category is direct, i.e. with a downwards-well-founded collection of non-identity morphisms, as is the case of semi-simplicial and semi-cubical sets, each $F(X)$ serving as source of a map $F(X) \stackrel{F(f)}{\rightarrow} F(Y)$ can be turned into a family $F(X)$ indexed on $F(Y)$, and, since the maps are well-founded downward, this can be iterated from below, eventually producing a collection of families which we call an \emph{indexed presentation} of the presheaf over the direct category.

@@

The idea for such indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types~\footnote{See e.g. \url{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}.}. A couple of constructions were proposed since then. One, by \cite{voevodsky12}, relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. Another, by \cite{herbelin15}\footnote{In hindsight, the title of the paper ``A dependently-typed construction of semi-simplicial types'' is somewhat confusing: it was part of a project for defining semi-simplicial types, but, in practice, the definition is done in a type theory with Unicity of Identity Proofs (UIP), meaning in a type theory where types (intended to represent arbitrary $\infty$-groupoids) are actually sets ($0$-truncated $\infty$-groupoids, that is, morally, $0$-groupoids). The confusion was however common at the time.}, and fully formalized in Coq, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by~\cite{part15}, formalized in (an emulation of logic-enriched homotopy type theory in) the Plastic proof assistant, and yet another by \cite{annenkovCK17}, formalized in (an emulation of a two-level type theory in) the Agda proof assistant~\footnote{\url{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, also rely on the presentation of the semi-simplicial category from increasing injective functions over finite ordinals.

The indexed definition of a presheaf over a direct category is technically more involved than the presheaf definition as it requires hard-wiring in the structure the dependencies between elements of the sets of the presheaf, including the coherence conditions between these dependencies, such as taking the $i$-th face of the $j$-th face of a $n$-simplex is the same as taking the $(j-1)$-th face of the $i$-th face (when $j>i$). On the other side, defining a concrete presheaf in indexed form only requires to provide the family since the responsibility of defining maps and showing the coherence conditions is already accounted for in the definition, what is not the case by default in the (ordinary) presheaf presentation.

In category-theoretic terms, the indexing set of the family $X_n$ in the indexed presentation is the \emph{matching object} of the presheaf at $n$. The indexed presentation of presheaves over a direct category can then be identified with a subclass of ordinary presheaves known as Reedy fibrant where Reedy fibrancy precisely expresses that the set $X_n$ in the fibered presentation is a $\Sigma$-type over an instance up to isomorphism of the matching object at level $n$~\citep{shulman14,kraus17}.

@@

While ordinary semi-simplicial sets are presheaves over the semi-simplex category, augmented semi-simplicial sets are presheaves over, what we will define as, \DeltaPlus. There are different ways to define \DeltaPlus, up to equivalence, and we use a definition that will later straightforwardly extend to semi-cubical sets.

@@

\section{Our construction}

From now on, we assume reasoning in type theory, first informally, then, stepwise, in afully formal type theory implements.u.

In informal type theory, which would roughly correspond to an extensional type theory leaving implicit trivial steps, our inductive construction is described by tables \ref{tab:coind}, \ref{tab:core}, \ref{tab:frames} and \ref{tab:faces}. However, we went to the point of fully formalizing it in the intentional type theory of the Coq proof assistant. Going to such level of details requires to address the following points:
\begin{itemize}
  \item Verification that the construction is well-typed: for instance, the typing of $\restrl$, given in table \ref{tab:faces}, requires a non-trivial coherence condition, given on~\ref{tab:coh}, which has to be proved simultaneously with the rest of the construction. The informal description of the whole construction, including the coherence condition, is explained in section~\ref{sec:ett}.
  \item The well-foundedness of the induction requires a special termination evidence which will be discussed in section~\ref{sec:wf}.
  \item The construction is indexed by integers and holds under some constraints on the range of these integers. There is a standard formalization dilemma in this kind of situation: either the constraints on the range are embedded in the construction so that the construction makes sense only on the corresponding range, or the construction is made first on a more general domain than needed but restricted to a smaller domain only at the time of use. We adopted the first approach, requiring however to let the construction be dependent on proofs of inequalities on natural numbers. We discuss how we deal with such dependencies in section~\ref{sec:le}.
  \item The interpretation of extensional type theory in intensional type theory requires to make explicit equality reasoning, leading to types of the construction becoming dependent on proofs of equalities. Our constructions also relies on a few standard type isomorphisms. Both issues will be discussed in Section~\ref{sec:itt}.
  \item Among those equalities, a number of groupoid properties of equality as well as type isomorphisms need to be made explicit. This will be discussed in section~\ref{sec:eqproperties}.
\end{itemize}

But first, let us give details and intuition on the informal construction.

@@

The definition is thus dispatched over tables \ref{tab:coind}, \ref{tab:core}, \ref{tab:frames} and \ref{tab:faces}. The extra table~\ref{tab:coh} shows the proof of a coherence condition use in \ref{tab:faces}. Table \ref{tab:coind} describes a $\nu$-set in indexed form, as a stream, coinductively representing the limit of $n$-truncated $\nu$-sets. The $n$-truncated $\nu$-sets are themselves described in \ref{tab:core}. In such stream, the $n$-th component is a type dependent over a $\fullframe$. The type $\fullframe$ is recursively defined in \ref{tab:frames}, using the auxiliary definitions of $\layer$ and $\painting$. The type $\layer$ and $\painting$ are dependent over $\restrf$, and these restrictions are defined on \ref{tab:faces}. These restrictions are defined using auxiliary definitions of $\restrl$ and $\restrc$.

@@

\renewcommandx{\framep}[5][1,2,3,4,5]{\prim{frame}[][#2][#3][#4][#5]}
\renewcommandx{\layer}[5][1,2,3,4,5]{\prim{layer}[][#2][#3][#4][#5]}
\renewcommandx{\painting}[5][1,2,3,4,5]{\prim{painting}[][#2][#3][#4][#5]}
\renewcommandx{\restrf}[7][1,2,3,4,5,6,7]{\restr{frame}[][#2][#3][#4][#5][#6][#7]}
\renewcommandx{\restrl}[7][1,2,3,4,5,6,7]{\restr{layer}[][#2][#3][#4][#5][#6][#7]}
\renewcommandx{\restrc}[7][1,2,3,4,5,6,7]{\restr{painting}[][#2][#3][#4][#5][#6][#7]}
\renewcommandx{\cohf}[9][1,2,3,4,5,6,7,8,9]{\coh{frame}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohl}[9][1,2,3,4,5,6,7,8,9]{\coh{layer}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohc}[9][1,2,3,4,5,6,7,8,9]{\coh{painting}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\renewcommand*{\lab}{fulltab:faces}
\input{tab-faces.tex}

>>>>

\section{Introduction}
\subsection*{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}
A family of sets can commonly be represented in two ways: as a family properly speaking, indexed over a given set $S$, or as a set $T$ together with a map specifying for each element of the set $T$ on which index in $S$ it depends. In the first case, we call it an \emph{indexed} presentation. In the latter case, the set associated to a given index is the fiber, i.e. inverse image, of the index under the map, so we call it a \emph{fibered} presentation. The two presentations are equivalent and the equivalence can be phrased concisely in the language of homotopy type theory~\citep{hottbook} as a fibered/indexed correspondence\footnote{At a discussion on the Zulip channel for the HoTT community, alternative names fibration/family correspondence and unbundled/bundled correspondence were proposed, among others. In our experience, the opposition fibered/indexed is spontaneously understood by many listeners. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories. The most elementary instance of the correspondence, with $\Type$ instead of $\U$, that is $(\Sigma S: \Type. (S \rightarrow T)) \simeq (T \rightarrow \Type)$, is sometimes called ``Grothendieck construction for dummies'' (its proof requires univalence, see \cite{hottbook}).}
\begin{equation*}
  \mbox{(fibered)}\qquad(\Sigma S: \U. (S \rightarrow T)) ~\simeq~ (T \rightarrow \U)\qquad \mbox{(indexed)}
\end{equation*}
where $\U$ represents in homotopy type theory the subset of types with a unique proof of equality (within some universe).

A \emph{presheaf} on an category is a family of sets connected with maps indexed over the morphisms of the category. As such, it lives on the indexed side of the correspondence, opposing on the fibered side to \emph{discrete Grothendieck fibrations}~\cite{FIND-A-CITATION}. However, there are situations where a presheaf $F$ can also be seen as living on the fibered side of the correspondence. Indeed, when the indexing category is \emph{direct}, i.e. with a downwards-well-founded collection of non-identity morphisms, as is the case of semi-simplicial and semi-cubical sets, each $F(X)$ serving as source of a map $F(X) \stackrel{F(f)}{\rightarrow} F(Y)$ can be turned into a family $F(X)$ indexed on $F(Y)$, and, since the maps are well-founded downward, this can be iterated from below, eventually producing a collection of families which we call an \emph{indexed presentation} of the presheaf over the direct category.

@@

The idea for such indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types~\footnote{See e.g. \url{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}.}. A couple of constructions were proposed since then. One, by \cite{voevodsky12}, relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. Another, by \cite{herbelin15}\footnote{In hindsight, the title of the paper ``A dependently-typed construction of semi-simplicial types'' is somewhat confusing: it was part of a project for defining semi-simplicial types, but, in practice, the definition is done in a type theory with Unicity of Identity Proofs (UIP), meaning in a type theory where types (intended to represent arbitrary $\infty$-groupoids) are actually sets ($0$-truncated $\infty$-groupoids, that is, morally, $0$-groupoids). The confusion between the question of moving from semi-simplicial sets to semi-simplicial types and the question of moving from a presheaf to an indexed presentation of it was however common at the time.}, and fully formalized in the Coq proof assistant, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by \cite{part15}, formalized in (an emulation of logic-enriched homotopy type theory in) the Plastic proof assistant, and yet another by \cite{annenkovCK17}, formalized in (an emulation of a two-level type theory in) the Agda proof assistant~\footnote{\url{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, also rely on the presentation of the semi-simplicial category from increasing injective functions over finite ordinals.

The indexed definition of a presheaf over a direct category is technically more involved than the presheaf definition as it requires hard-wiring in the structure the dependencies between elements of the sets of the presheaf, including the coherence conditions between these dependencies, such as taking the $i$-th face of the $j$-th face of a $n$-simplex being the same as taking the $(j-1)$-th face of the $i$-th face (when $j>i$). On the other side, defining a concrete presheaf in indexed form only requires to provide the family since the responsibility of defining maps and showing the coherence conditions is already accounted for in the definition, what is not the case by default in the (ordinary) presheaf presentation.

In category-theoretic terms, the indexing set of the family $X_n$ in the indexed presentation is a \emph{matching object} of the presheaf at $n$. The indexed presentation of presheaves over a direct category can then be identified with a subclass of ordinary presheaves known as Reedy fibrant where Reedy fibrancy precisely expresses that the set $X_n$ in the fibered presentation is a $\Sigma$-type over a representative of the class of equivalence of matching objects at level $n$~\citep{shulman14,kraus17}.

@@

While ordinary semi-simplicial sets are presheaves over the semi-simplex category, augmented semi-simplicial sets are presheaves over, what we will define as, \DeltaPlus. There are different ways to define \DeltaPlus, up to equivalence, and we use a definition that can be extended to semi-cubical sets in a straightforward manner.

@@

\section{Our construction}
In this section, we describe our parametricity-based construction of $\nu$-sets in indexed form at different levels of formality:

\begin{itemize}
  \item Sections~\ref{sec:ett} and~\ref{sec:itt} describe the construction at an informal level of discourse:
        \begin{itemize}
          \item In \ref{sec:ett}, we present it in informal extensional type theory where equational reasoning is left implicit.
          \item While reasoning in extensional type theory is similar to reasoning in set theory regarding how equality is implicitly handled, extensional type theory has two limitations. The first limitation is that it enforces the principle of Uniqueness of Identity Proofs and this is inconsistent with the Univalence principle. The second limitation is that we formalized the construction in the Coq proof assistant whose underlying type theory is intensional. Section~\ref{sec:itt} thus rephrases the construction in (informal) intensional type theory.
        \end{itemize}
  \item Sections~\ref{sec:wf}, \ref{sec:le} and \ref{sec:eqproperties} describe issues to be addressed in order to get a fully formal construction:
        \begin{itemize}
          \item The well-foundedness of the induction requires a special termination evidence which will be discussed in section~\ref{sec:wf}.
          \item The construction is indexed over integers and holds under some constraints on the range of these integers. There is a standard formalization dilemma in this kind of situation: either the constraints on the range are embedded in the construction so that the construction makes sense only on the corresponding range, or the construction is made first on a more general domain than needed but restricted to a smaller domain only at the time of use. We adopted the first approach, requiring however to let the construction be dependent on proofs of inequalities on natural numbers. We discuss how we deal with such dependencies in section~\ref{sec:le}.
          \item A number of obvious groupoid properties of equality as well as type isomorphisms have been left implicit in the informal definition. This is discussed in section~\ref{sec:eqproperties}.
        \end{itemize}
\end{itemize}

@@

The definition is dispatched over tables \ref{tab:coind}, \ref{tab:core}, \ref{tab:frames} and \ref{tab:faces}. The extra table~\ref{tab:coh} shows the proof of a coherence condition use in \ref{tab:faces}. Table \ref{tab:coind} describes a $\nu$-set in indexed form, as a coinductively-defined infinite sequence of types representing the limit over $n$ of a $n$-truncated $\nu$-set. The definition of a $n$-truncated $\nu$-set is in turn described in \ref{tab:core}. In the infinite sequence of types representing a $\nu$-set, the $n$-th component is a type dependent over a $\fullframe$. The type $\fullframe$ is recursively defined in \ref{tab:frames}, using the auxiliary definitions of $\layer$ and $\painting$. The type $\layer$ and $\painting$ are dependent over $\restrf$, and these restrictions are defined on \ref{tab:faces}. These restrictions are defined using auxiliary definitions of $\restrl$ and $\restrc$.

@@

% Insert fulltab:faces and fulltab:coh, after overriding macros to make
% everything but the universe letter explicit.
\renewcommandx{\framep}[5][1,2,3,4,5]{\prim{frame}[][#2][#3][#4][#5]}
\renewcommandx{\layer}[5][1,2,3,4,5]{\prim{layer}[][#2][#3][#4][#5]}
\renewcommandx{\painting}[5][1,2,3,4,5]{\prim{painting}[][#2][#3][#4][#5]}
\renewcommandx{\restrf}[7][1,2,3,4,5,6,7]{\restr{frame}[][#2][#3][#4][#5][#6][#7]}
\renewcommandx{\restrl}[7][1,2,3,4,5,6,7]{\restr{layer}[][#2][#3][#4][#5][#6][#7]}
\renewcommandx{\restrc}[7][1,2,3,4,5,6,7]{\restr{painting}[][#2][#3][#4][#5][#6][#7]}
\renewcommandx{\cohf}[9][1,2,3,4,5,6,7,8,9]{\coh{frame}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohl}[9][1,2,3,4,5,6,7,8,9]{\coh{layer}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohc}[9][1,2,3,4,5,6,7,8,9]{\coh{painting}[][#2][#3][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\coht}[9][1,2,3,4,5,6,7,8,9]{\cohtwo{frame}[][#2][#3][#4][#5][#6][#7,#8][#9]}
\renewcommand*{\lab}{fulltab:faces}
\input{tab-faces.tex}
\renewcommand*{\lab}{fulltab:coh}
\input{tab-coh.tex}
